---
title: "MDVC optimization code"
author: "B. Raynor, A. Saxena"
date: "12/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code accompanies the manuscript:
Castillo et al, Optimizing the location of vaccination sites to stop a zoonotic epidemic

## Load and format data  

**Data availability:** Data used for this analysis includes personally identifiable information (the geographic location of homes). For this reason, data files will not be made publically available. For questions or to request access to this data contract corresponding author, Ricardo Castillo (cricardo@upenn.edu).  

Load packages  
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tbart)
library(lme4)
library(ggmap)
library(ggsn)
library(sp)
library(rgdal)
```  

Load and format house data
```{r, warning=FALSE, message=FALSE}
setwd("~/Rabies")

#Function to convert to utm
UTMfunction <- function(lat, long) {
  coord <- data.frame(X=long, Y=lat)
  df.sp.latlong <- SpatialPoints(coord, proj4string=CRS("+proj=longlat +ellps=WGS84"))
  df.sp.utm <- spTransform(df.sp.latlong, CRS("+proj=utm +south +zone=19K  +ellps=WGS84"))
  UTMcoord <- as.data.frame(coordinates(df.sp.utm))
  return(UTMcoord)
}

#load in information about vaccinations strategy(whether fixed, mobile, or mixed)
vac_strat<-read.csv("~/Rabies/VANCAN/Vaccination Strategy.csv") %>%
  mutate(filter= sub("1.1.", "", Loc_code))%>% #Identify non unique
  mutate(filter=as.numeric(filter))%>% #unique should all be numberic
  filter(!is.na(filter))%>% #filter out non numeric
  select(-X,-X.1,-filter)%>% #remove useless columns
  mutate(L=c(1:length(Loc_name))) #create ID column


#load in house data 2016 - 2018
all_house_data<-read.csv("~/Rabies/rabia_ASA_encuesta_2018/resultados/houses_merged_2019-02-18_vac.csv") %>%
  mutate(X = UTMfunction(LATITUDE, LONGITUDE)$X, 
         Y = UTMfunction(LATITUDE, LONGITUDE)$Y)
all_house_data<-dplyr::full_join(all_house_data,vac_strat,by='L') #merge vacc strategies

#subset out years
house_2016_data <- all_house_data %>%
  filter(year2016==1)%>%
  select(UNICODE, LATITUDE, LONGITUDE, X, Y, VIVIENDA_VANCAN_2016, Strategy_16)%>%
  filter(Strategy_16 == "F")%>%
  select(-Strategy_16)

house_2017_data <- all_house_data %>%
  filter(year2017==1)%>%
  select(UNICODE, LATITUDE, LONGITUDE, X, Y, VIVIENDA_VANCAN_2017, Strategy_17)%>%
  filter(Strategy_17 == "F")%>%
#  filter(Strategy_17 != "M")%>%
  select(-Strategy_17)

house_2018_data <- all_house_data %>%
  filter(year2018==1)%>%
  select(UNICODE, LATITUDE, LONGITUDE, X, Y, VIVIENDA_VANCAN_2018, Strategy_18)%>%
  filter(Strategy_18 == "F")%>%
#  filter(Strategy_18 != "M")%>%
  select(-Strategy_18)

house_2019_data <- left_join(
  read.csv('~/Rabies/rabia_ASA_encuesta_2019/resultados/VIVIENDA2019DCE_CLEAN_2019-11-18.csv')%>%
    rename(UNICODE=VIVIENDA_UNICODE),
  read.csv('~/Rabies/shortest_way_optimization/demandpoints/demand_points.csv', sep=";")%>%
    select(UNICODE,lat,long))%>%
  tidyr::drop_na(lat)%>% ####CHECK THIS 
  mutate(X = UTMfunction(lat, long)$X, 
         Y = UTMfunction(lat, long)$Y) %>%
  rename(LATITUDE = lat,
         LONGITUDE = long)%>%
  select(UNICODE, LATITUDE, LONGITUDE, X, Y, VIVIENDA_HAY_PERROS, 
         VIVIENDA_DONDE_VACUNARON) %>%
  filter(VIVIENDA_HAY_PERROS==TRUE)%>%
  mutate(VIVIENDA_VANCAN_2019 = ifelse(is.na(VIVIENDA_DONDE_VACUNARON), 0, 1))%>%
  select(-VIVIENDA_HAY_PERROS, -VIVIENDA_DONDE_VACUNARON)

```

Load and format vaccination point data
```{r, warning=FALSE, message=FALSE}
#General
vac_points_gen <- read.csv("~/Rabies/VANCAN/VaccinationPoints/PUNTOS_VANCAN_General_CURRENT.csv")

#Each year
vac_points_2016<-read.csv("~/Rabies/VANCAN/VaccinationPoints/PUNTOS_VANCAN_2016_Indexed_2021.08.26.csv")
vac_points_2016 <- left_join(vac_points_2016, vac_points_gen) %>%
  filter(TypeFix == "F")%>%
  mutate(X = UTMfunction(lat, long)$X,
         Y = UTMfunction(lat, long)$Y)%>%
  select(VaccPoint, PointName, yearlabel, lat, long, X, Y, TypeFix)


#2017
vac_points_2017<-read.csv("~/Rabies/VANCAN/VaccinationPoints/PUNTOS_VANCAN_2017_Indexed_2021.08.26.csv")
vac_points_2017 <- left_join(vac_points_2017, vac_points_gen)%>%
  filter(TypeFix == "F")%>%
  mutate(X = UTMfunction(lat, long)$X,
         Y = UTMfunction(lat, long)$Y)%>%
  select(VaccPoint, PointName, yearlabel, lat, long, X, Y, TypeFix)

#2018
vac_points_2018<-read.csv('~/Rabies/rabia_ASA_encuesta_2018/base/Puntos_VANCAN_General.csv',sep=';')%>%
  select(VaccPoint)%>%
  mutate(yearlabel=2018)
vac_points_2018 <- left_join(vac_points_2018, vac_points_gen)%>%
  filter(TypeFix == "F")%>%
  mutate(X = UTMfunction(lat, long)$X,
         Y = UTMfunction(lat, long)$Y)%>%
  select(VaccPoint, PointName, yearlabel, lat, long, X, Y, TypeFix)


#2019
vac_points_2019<-read.csv("~/Rabies/VANCAN/VaccinationPoints/PUNTOS_VANCAN_2019_Indexed_2021.08.26.csv")
vac_points_2019 <- left_join(vac_points_2019, vac_points_gen)%>%
  filter(TypeFix == "F")%>%
  mutate(X = UTMfunction(lat, long)$X,
         Y = UTMfunction(lat, long)$Y)%>%
  select(VaccPoint, PointName, yearlabel, lat, long, X, Y, TypeFix)
```

Map of study area
**Fig 1. Alto Selva Alegre district with possible vaccination locations**
```{r, warning=FALSE, message=FALSE}
#2016 vancan
# Set projection
latlon_CRS <- CRS("+proj=longlat +ellps=WGS84")

#pull district shapefle and format
limits <- rgdal::readOGR("~/Rabies/Shapefiles AQP/db/InstitutoGeograficoNacional/distritos/DISTRITOS.shp")
limits_shp.new <- spTransform(limits,latlon_CRS)
ASA<- limits_shp.new[limits_shp.new$DISTRITO %in% c("ALTO SELVA ALEGRE"),]

#basemap
basemap <- ggmap(get_stamenmap(maptype='terrain',bbox=c(left=-71.54,right=-71.488,top=-16.332,bottom=-16.392),zoom=14)) 

#add layers
ASAmap <-basemap +
  geom_polygon(data =ASA, aes(x = long, y = lat, group = group), 
                 colour = "black", size=1, fill = "black", alpha=0.3)+
  geom_point(data=vac_points_gen, aes(x=long,y=lat),fill="white",shape=24,size=3)+
  theme(legend.text=element_text(size=5))+
  scalebar(x.min = -71.53, x.max = -71.495,  y.min = -16.39, y.max = -16.34, 
           dist = 1, dist_unit = "km", st.bottom = FALSE, st.color = "black",
           st.dist=0.05, transform = TRUE, model = "WGS84")+
  ggtitle('Alto Selva Alegra District')+
  theme(legend.position=c(0.85,0.85),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        rect = element_blank(),
        legend.text=element_text(size=8,face='bold'))
  north2(ASAmap, scale=0.12, x=.65,y=.25)
```

Load and format walking distance matrix
```{r, warning=FALSE, message=FALSE}
################################################################################
#Load distance matrices
################################################################################

setwd("~/Rabies/shortest_way_optimization/results")

#original results
wd0 <- rbind(read.csv('ASA_SWD1_2021.08.05.csv'),
            read.csv('ASA_SWD2_2021.08.05.csv'),
            read.csv('ASA_SWD3_2021.08.05.csv'),
            read.csv('ASA_SWD4_2021.08.05.csv'),
            read.csv('ASA_SWD5_2021.08.05.csv'),
            read.csv('ASA_SWD6_2021.08.05.csv'),
            read.csv('ASA_SWD7_2021.08.05.csv'),
            read.csv('ASA_SWD8_2021.08.05.csv'),
            read.csv('ASA_SWD9_2021.08.05.csv'))%>%
  select(UNICODE_house, VaccPoint, PointName, distancia) %>%
  rename(UNICODE = UNICODE_house)

#new results calculated by Gian Sept 2021
wd1 <- rbind(read.csv('result_Puntos_VANCAN_General_NeedWalkingDistance_1.csv'),
             read.csv('result_Puntos_VANCAN_General_NeedWalkingDistance_2.csv'),
             read.csv('result_Puntos_VANCAN_General_NeedWalkingDistance_3.csv'),
             read.csv('result_Puntos_VANCAN_General_NeedWalkingDistance_4.csv'),
             read.csv('result_Puntos_VANCAN_General_CURRENT_point3.csv'),
             read.csv('result_Puntos_VANCAN_General_CURRENT_point2_1.csv'),
             read.csv('result_Puntos_VANCAN_General_CURRENT_point2_2.csv'))

#bind walking distance results together         
wd <- bind_rows(wd0, wd1)

################################################################################
#Format distance matrices
################################################################################
#2016 distance matrix
dist.2016<-wd%>% #filter main walking distance matrix for vacc points and houses in 2016
  filter(VaccPoint %in% vac_points_2016$VaccPoint)%>% #vacc points
  filter(UNICODE %in% house_2016_data$UNICODE)%>%#house
  select(VaccPoint, UNICODE, distancia)%>%
  tidyr::pivot_wider(names_from =VaccPoint, values_from = distancia)%>%
  tibble::column_to_rownames('UNICODE')

#2017 distance matrix
dist.2017<-wd%>% #filter main walking distance matrix for vacc points and houses in 2016
  filter(VaccPoint %in% vac_points_2017$VaccPoint)%>% #vacc points
  filter(UNICODE %in% house_2017_data$UNICODE)%>%#house
  select(VaccPoint, UNICODE, distancia)%>%
  tidyr::pivot_wider(names_from =VaccPoint, values_from = distancia)%>%
  tibble::column_to_rownames('UNICODE')

#2018 distance matrix
dist.2018<-wd%>% #filter main walking distance matrix for vacc points and houses in 2016
  filter(VaccPoint %in% vac_points_2018$VaccPoint)%>% #vacc points
  filter(UNICODE %in% house_2018_data$UNICODE)%>%#house
  select(VaccPoint, UNICODE, distancia)%>%
  tidyr::pivot_wider(names_from =VaccPoint, values_from = distancia)%>%
  tibble::column_to_rownames('UNICODE')

#2019 distance matrix
dist.2019<-wd%>% #filter main walking distance matrix for vacc points and houses in 2016
  filter(VaccPoint %in% vac_points_2019$VaccPoint)%>% #vacc points
  filter(UNICODE %in% house_2019_data$UNICODE)%>%#house
  select(VaccPoint, UNICODE, distancia)%>%
  tidyr::pivot_wider(names_from =VaccPoint, values_from = distancia)%>%
  tibble::column_to_rownames('UNICODE')


#Function to find closest vaccination point for every house
ClosestVaccFunction <- function(distance.matrix, min.matrix){
  for(i in 1:nrow(distance.matrix)){
    houseID <- rownames(distance.matrix)
    vacID <- colnames(distance.matrix)
    
    #Find minimum distance
    house <- distance.matrix[i,]
    min.dist <- min(house)
  
    #Index house and vacc point for min distance
    houseIndex <- houseID[i]
    vacIndex <- which(house == min.dist)[[1]]
    vacIndex <- vacID[vacIndex]
    
    #save to matrix
    min.matrix <- rbind(min.matrix, c(houseIndex, vacIndex, min.dist))
  }
  min.matrix <- as.data.frame(min.matrix)
  colnames(min.matrix) <- c("UNICODE", "Vacc_ID", "distance")
  return(min.matrix)
}

#Find closest vaccination point based on minimum distance in distance matrix

#2016
newmins <-NULL
newmins <- ClosestVaccFunction(dist.2016, newmins)
house_2016_data <-full_join(house_2016_data, newmins)%>%
  rename(lat = LATITUDE,
         long =LONGITUDE,
         vac_status = VIVIENDA_VANCAN_2016) %>%
  mutate(year=2016)

#2017
newmins <-NULL
newmins <- ClosestVaccFunction(dist.2017, newmins)
house_2017_data <-full_join(house_2017_data, newmins)%>%
  rename(lat = LATITUDE,
         long =LONGITUDE,
         vac_status = VIVIENDA_VANCAN_2017) %>%
  mutate(year=2017)

#2018
newmins <-NULL
newmins <- ClosestVaccFunction(dist.2018, newmins)
house_2018_data <-full_join(house_2018_data, newmins)%>%
  rename(lat = LATITUDE,
         long =LONGITUDE,
         vac_status = VIVIENDA_VANCAN_2018) %>%
  mutate(year=2018)

#2019
newmins <-NULL
newmins <- ClosestVaccFunction(dist.2019, newmins)
house_2019_data <-full_join(house_2019_data, newmins)%>%
  rename(lat = LATITUDE,
         long =LONGITUDE,
         vac_status = VIVIENDA_VANCAN_2019) %>%
  mutate(year=2019)

#bind together into one dataframe
vac_data <- bind_rows(house_2016_data, 
                      house_2017_data, 
                      house_2018_data, 
                      house_2019_data, 
                      )
vac_data$distance <- as.numeric(vac_data$distance)
```

### Participation probability regression

# Regressions
Create bins of data for count regressions
```{r}
vac_data <- vac_data%>%
  mutate(Bin = NA)

#Create bins
BinWidth = 150 #THIS IS THE SIZE OF THE BINS
MaxSeq <- ceiling(max(vac_data$distance)/BinWidth) #autoset to cover all distances
bins <- data.frame(BinNum = seq(1:MaxSeq),
                   StartDist= BinWidth*(seq(1:MaxSeq) -1),
                   EndDist = BinWidth*seq(1:MaxSeq)) #ref table


#Assign a bin number based on ref table
for(i in 1:nrow(bins)){
  row = bins[i,]
  
  #training data
  vac_data <- vac_data %>%
    mutate(Bin = ifelse(
      distance >= row$StartDist[[1]] & distance < row$EndDist[[1]],row$BinNum[[1]], Bin))
}


#bin the training set
vac_data.bin <- data.frame(plyr::count(vac_data%>%select(vac_status, Bin, year)))%>%
  rename(BinNum = Bin)%>%
  tidyr::pivot_wider(names_from=vac_status, values_from=freq)%>%
  rename(num_not_vac = '0',
         num_vac ='1')%>%
  mutate(num_vac = tidyr::replace_na(num_vac, 0),
         num_not_vac = tidyr::replace_na(num_not_vac, 0),
         num_of_houses = num_not_vac+num_vac,
         VaccFreq = num_vac/num_of_houses)

vac_data.bin <- left_join(vac_data.bin, bins) %>% #add distance refs
  mutate(mean_dist=(StartDist + EndDist)/2)

vac_data_pois <- vac_data.bin
```

Run regressions
```{r}
#poisson
poisreg<-glm(num_vac~mean_dist+offset(log(num_of_houses)),
             family=poisson, data=vac_data_pois)
#gaussian
gaussreg<-glm(num_vac~mean_dist+offset(log(num_of_houses)),
             family=gaussian, data=vac_data_pois)

#negative binomial
neg_binom<-MASS::glm.nb(num_vac~mean_dist+offset(log(num_of_houses)), data=vac_data_pois)

#binomial
model_binom<-glm(formula = vac_status ~ distance, family = "binomial", 
               data = vac_data)

```

View table of results
```{r}
table <- data.frame(
  model = c("poisson", "gaussian", "negative binomial", "binomial"),
  formula = c(paste0(summary(poisreg)$call[2]), paste0(summary(gaussreg)$call[2]), 
              paste0(summary(neg_binom)$call[2]), paste0(summary(model_binom)$call[2])),
  intercept_beta = c(summary(poisreg)$coefficients[[1,1]], summary(gaussreg)$coefficients[[1,1]], 
                summary(neg_binom)$coefficients[[1,1]], summary(model_binom)$coefficients[[1,1]]),
  intercept_pval = c(summary(poisreg)$coefficients[[1,4]], summary(gaussreg)$coefficients[[1,4]], 
                summary(neg_binom)$coefficients[[1,4]], summary(model_binom)$coefficients[[1,4]]),
  distance_beta = c(summary(poisreg)$coefficients[[2,1]], summary(gaussreg)$coefficients[[2,1]], 
                summary(neg_binom)$coefficients[[2,1]], summary(model_binom)$coefficients[[2,1]]),
  distance_pval = c(summary(poisreg)$coefficients[[2,4]], summary(gaussreg)$coefficients[[2,4]], 
                summary(neg_binom)$coefficients[[2,4]], summary(model_binom)$coefficients[[2,4]]),
  AIC = c(summary(poisreg)$aic, summary(gaussreg)$aic, summary(neg_binom)$aic, summary(model_binom)$aic)
  )

table
```

Figure 2
```{r}
#check
ggplot()+
  theme_classic()+
  geom_point(data=vac_data.bin, aes(x=mean_dist, y=VaccFreq, size=num_of_houses, color=as.factor(year)))+
  ggtitle("Binned data")+
  xlab("Bin mid distance") + ylab("Vaccinated Frequency")
```


## Tbart Optimization